Re-examining cross-cultural similarity judgments using language statistics

### 2023-02-02 code/4_paper/manuscript/manuscript.Rmd
italics

An influential perspective within cultural psychology links these differences in similarity judgment to tendencies toward analytic processing in Western cultures, and holistic processing in East Asian cultures [@Nisbett2003].
Analytic processing emphasizes rule-like relationships predicated on objects and their properties and, correspondingly, taxonomic similarity, while holistic processing emphasizes relations between objects and their context and therefor, thematic similarity.
In related work, East Asian participants show a higher level of sensitivity to context than their Western counterparts when reproducing drawings from memory [@Ji2000]; visually exploring naturalistic scenes [@Chua2005]; describing scenes [@Masuda2001]; and in explaining the causes of ambiguous behaviors [@Choi1999]. The consensus interpretation in this literature ascribes cross-cultural differences in similarity judgment to variation in the *conceptualization* of similarity -- with people from East Asian cultures relying on a more thematic notion of similarity than Westerners.

@@

Language statistics may incorporate both cultural-specific environmental statistics, that is, the experiential *input* to similarity judgments, and culture-specific senses of similarity, the *conceptualization* of similarity.
For example, cultural features (like farming) can lead to differences in environmental statistics (seeing cows and grass) and these can influence language (talking more about cows and grass).
But other cultural features (like conceiving of similarity thematically) could also cause individuals to talk differently about the same experiences (mentioning what cows eat rather than what other animals cows are like).
Accordingly, our approach examines the extent to which language statistics can predict cross-cultural differences in similarity judgments with the understanding that language statistics are likely a proxy for both *input* to and *conceptualization* of similarity.

@@

To preview our results, we replicate US-China differences and find that Vietnamese judgments are intermediate between these two. We find that language-specific statistics provide good predictions for cross-cultural differences in similarity judgments, and additionally, for more general similarity judgments that do not contrast taxonomic and thematic matches. Critically, the stimuli used to assess this more general case of similarity reasoning were constructed to be outside the scope of explanation for taxonomic-thematic accounts (for use as filler items). While the language model may succeed in taxonomic-thematic similarity predictions by picking up on differences in the *conceptualization* of similarity that are reflected in language data, this conceptualization account provides no prediction for these filler stimuli. 
<!-- Accordingly, these filler stimuli provide the strongest test of the language model, demonstrating that it can explain variation in similarity judgments beyond that explained by thematic or taxonomic conceptualizations of similarity.  -->

>>>>

An influential perspective within cultural psychology links these differences in similarity judgment to tendencies toward analytic processing in Western cultures, and holistic processing in East Asian cultures [@Nisbett2003].
Analytic processing emphasizes rule-like relationships predicated on objects and their properties and, correspondingly, taxonomic similarity, while holistic processing emphasizes relations between objects and their context and therefor, thematic similarity.
In related work, East Asian participants show a higher level of sensitivity to context than their Western counterparts when reproducing drawings from memory [@Ji2000]; visually exploring naturalistic scenes [@Chua2005]; describing scenes [@Masuda2001]; and in explaining the causes of ambiguous behaviors [@Choi1999]. The consensus interpretation in this literature ascribes cross-cultural differences in similarity judgment to variation in the conceptualization of similarity -- with people from East Asian cultures relying on a more thematic notion of similarity than Westerners.

@@

Language statistics may incorporate both cultural-specific environmental statistics, that is, the experiential *input* to similarity judgments, and culture-specific senses of similarity, the *conceptualization* of similarity.
For example, cultural features (like farming) can lead to differences in environmental statistics (seeing cows and grass) and these can influence language (talking more about cows and grass).
But other cultural features (like conceiving of similarity thematically) could also cause individuals to talk differently about the same experiences (mentioning what cows eat rather than what other animals cows are like).
Accordingly, our approach examines the extent to which language statistics can predict cross-cultural differences in similarity judgments with the understanding that language statistics are likely a proxy for both input and conceptualization.

@@

To preview our results, we replicate US-China differences and find that Vietnamese judgments are intermediate between these two. We find that language-specific statistics provide good predictions for cross-cultural differences in similarity judgments, and additionally, for more general similarity judgments that do not contrast taxonomic and thematic matches. Critically, the stimuli used to assess this more general case of similarity reasoning were constructed to be outside the scope of explanation for taxonomic-thematic accounts (for use as filler items). While the language model may succeed in taxonomic-thematic similarity predictions by picking up on differences in the conceptualization of similarity that are reflected in language data, this conceptualization account provides no prediction for these filler stimuli. 
<!-- Accordingly, these filler stimuli provide the strongest test of the language model, demonstrating that it can explain variation in similarity judgments beyond that explained by thematic or taxonomic conceptualizations of similarity.  -->
